### 2019-12-16 ic-journal/coarsest.tex
Fixed other small issues per email discussions.

To prove Theorem~\ref{thm:coarsestfiniteState}, it only remains to construct such stable
process $k$ for any two finite-state processes $p$ and $q$.
For arbitrary CCS processes, this construction relies on
arbitrary infinite sums of processes (not within our CCS syntax) and
transfinite induction to obtain
an arbitrary large sequence of processes that are all pairwise
non-bisimilar,
which was firstly introduced by Jan
Willem Klop~(see \cite{van2005characterisation} for some historical notes).
We have only partially formalised
van Glabbeek's proof, mostly because our CCS syntax does not allow infinite
summation (and it is not easy to extend it with this support).
Another more important reason is that, the typed logic
implemented in various HOL systems (including Isabelle/HOL) is not
strong enough to define a type for all possible
ordinals~\cite{norrish2013ordinals} which is required in van
Glabbeek's proof. As the consequence, the formalisation
(Theorem~\ref{thm:coarsestfiniteState})
can only apply to finite-state CCS.

>>>>

To prove Theorem~\ref{thm:coarsestfiniteState}, it only remains to construct such stable
process $k$ for any two finite-state processes $p$ and $q$.
For arbitrary CCS processes, this construction relies on
arbitrary infinite sums of processes (not within our CCS syntax) and
transfinite induction to obtain
an arbitrary large sequence of processes that are all pairwise
non-bisimilar,
which was firstly introduced by Jan
Willem Klop~(see \cite{van2005characterisation} for some historical notes).
We have only partially formalised
van Glabbeek's proof, mostly because our CCS syntax does not allow infinite
summation (and it is not easy to extend it with this support).
Another more important reason is that the typed logic
implemented in various HOL systems (including Isabelle/HOL) is not
strong enough to define a type for all possible
ordinals~\cite{norrish2013ordinals} which is required in van
Glabbeek's proof. As the consequence, the formalisation
(Theorem~\ref{thm:coarsestfiniteState})
can only apply to finite-state CCS.

### 2019-12-16 ic-journal/multivariate.tex
Fixed other small issues per email discussions.

In the literature (\citep[p.~102]{Gorrieri:2015jt}, e.g.), the variables $X_i$ of
a system of equations $\{X_i = E_i\}_{i\in I}$ are usually considered as
\emph{process variables} outside the CCS syntax.
Then, an equation body  $E_i$  is an \emph{open} expression
built with CCS operators plus these equation variables.
In a  formalisation, this means either  defining a
whole new datatype (with extra equation variables), in which each CCS
operator must be duplicated, or adding equation variables as a
new primitive (e.g.~\texttt{Var X}) to the existing CCS type (besides
 agent variables such as \texttt{var X}).
In either cases the \emph{disjointness} between equation variables and
agent variables is syntactically
guaranteed. Both solutions are rather combersome, and require a non-trivial modification
of the existing \univariate formalisation.

@@

As mentioned at the beginning of Section~\ref{s:eq},
within our CCS syntax an agent variable may occur outside the recursion in which it is bound.
Thus the same variable can appear both \emph{free} and \emph{bound} in a CCS term.
%
We denote the set of bound variables of a
given CCS term $E$ (the variables that are bound in a recursion
subexpression of $E$) as $\bv{E}$ (or \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}} in HOL), and the set of  free
variables as $\fv{E}$ (or \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}} in HOL). Both \HOLinline{\HOLConst{BV}} and \HOLinline{\HOLConst{FV}} have the
type ``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{\alpha} \HOLTokenTransEnd \HOLTyOp{bool}}'', i.e. functions taking CCS terms and returning
(finite) sets of variables (of the type \HOLinline{\ensuremath{\alpha}}).
% Milner didn't give their definitions explicitly. However, one can
% imagine that, for most CCS operators they just pass into
% subterms without collecting anything (and $\nil$ gives $\emptyset$).
For their definition the interesting cases are those of
recursion and agent variables, shown below  (here \texttt{DELETE}
and \texttt{INSERT} are set-theoretic operators of HOL's
\texttt{pred_set} theory):
\begin{center}
\begin{tabular}{|l|l|}
\hline
\HOLConst{FV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLTokenLeftbrace{}\HOLFreeVar{X}\HOLTokenRightbrace{} & \HOLConst{FV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLTokenDefEquality{} \HOLConst{FV} \HOLFreeVar{p} \HOLConst{DELETE} \HOLFreeVar{X} \\
\HOLConst{BV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenEmpty{}} & \HOLConst{BV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLTokenDefEquality{} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{p} \\
\hline
\end{tabular}
\end{center}
Furthermore, $E$ is a process, written \HOLinline{\HOLConst{IS_PROC} \HOLFreeVar{E}}, if
it does not contain any free variable, i.e. $\fv{E} = \emptyset$:
\begin{alltt}
   \HOLConst{IS_PROC} \HOLFreeVar{E} \HOLTokenDefEquality{} \HOLConst{FV} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\HOLTokenEmpty{}}\hfill{[IS_PROC_def]}
\end{alltt}
And a list of CCS processes can be asserted by \HOLinline{\HOLConst{ALL_PROC}} defined upon \HOLinline{\HOLConst{IS_PROC}}:
\begin{alltt}
   \HOLConst{ALL_PROC} \HOLFreeVar{Es} \HOLTokenDefEquality{} \HOLConst{EVERY} \HOLConst{IS_PROC} \HOLFreeVar{Es}\hfill{[ALL_PROC_def]}
\end{alltt}

The set of free and bound
variables of a term need not be disjoint, e.g.~in $X + \recu X E$.
More importantly, when going
from a CCS expression to its sub-expressions, the set of free
variables may increase, while the set of bound variables either
remains the same or decreases. For instance, $\fv{\recu X (\mu.X)} = \emptyset$, while
$\fv{\mu.X} = \{X\}$. This property of $\fv{\cdot}$ brings some
difficulities in transition inductions. As an evidence,
we prove the following fundamental property of
$\fv{\cdot}$~\citep[p.~1209]{milner1990operational}:
%
\begin{proposition}
\label{prop:transFV}
The derivatives of a process are themselves processes, i.e.
if $E \arr{\mu} E'$ and $\fv{E} = \emptyset$, then $\fv{E'} =
\emptyset$. Formally:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{IS_PROC} \HOLBoundVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLConst{IS_PROC} \ensuremath{\HOLBoundVar{E}\sp{\prime}}\hfill{[TRANS_PROC]}
\end{alltt}
\end{proposition}

@@

As a matter of fact, in the HOL proof ``action induction''

@@

Note that \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}\\\;\HOLConst{DELETE}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{=}}\\\;\HOLConst{FV}\\\;\ensuremath{(}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{)}}, so
this is indeed the consequence of action induction on the Recursion.

@@

Proposition~\ref{prop:transFV} is essential in the proofs of the
  \multivariate versions of all unique-solution theorems.
On the other hand, the analogous
result for bound variables is indeed just an easy transition
induction. (The easy proof is omitted.)
\begin{proposition}
\label{prop:transBV}
if $E \arr{\mu} E'$ then $\bv{E'} \subseteq \bv{E}$, or formally:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{BV} \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLBoundVar{E}\hfill{[TRANS_BV]}
\end{alltt}
\end{proposition}
% \begin{proof}
% Following the proof of Proposition~\ref{prop:transFV}, we do action
% induction on $E \arr{\mu} E'$ and simplify all subgoals by set-theoretic
% facts. The only subgoal left is:
% \begin{alltt}
%         \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{E}}
%    ------------------------------------
%     0.  \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}}
% \end{alltt}
% Note the difference with the proof of Proposition~\ref{prop:transFV}: now
% we have \texttt{INSERT} in place of \texttt{DELETE}. Then, we directly
% use the assumption and the transitivity of $\subseteq$ and try to
% prove the following goal instead:
% \begin{alltt}
%         \HOLinline{\HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{E}}
%    ------------------------------------
%     0.  \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}}
% \end{alltt}
% This is an easy goal, due to the following lemma which is
%  proved by doing induction on $E$:
% \begin{alltt}
% \HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{X} \HOLBoundVar{E} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLConst{BV} \ensuremath{(}\ensuremath{[}\ensuremath{\HOLBoundVar{E}\sp{\prime}}\ensuremath{/}\HOLBoundVar{X}\ensuremath{]} \HOLBoundVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLBoundVar{E} \HOLSymConst{\HOLTokenUnion{}} \HOLConst{BV} \ensuremath{\HOLBoundVar{E}\sp{\prime}}\hfill{[BV_SUBSET]}
% \end{alltt}
% Taking $E ' = \recu X E$ in the above lemma, we have
%     \HOLinline{\HOLConst{BV}\\\;\ensuremath{(}\ensuremath{[}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]}\\\;\HOLFreeVar{E}\ensuremath{)}\\\;\HOLSymConst{\HOLTokenSubset{}}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}\\\;\HOLSymConst{\HOLTokenUnion{}}\\\;\HOLConst{BV}\\\;\ensuremath{(}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{)}},
%     the right side equals to \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}\\\;\HOLSymConst{\HOLTokenUnion{}}\\\;\ensuremath{(}\HOLFreeVar{X}\\\;\HOLConst{INSERT}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}\ensuremath{)}} $=$
%     \HOLinline{\HOLFreeVar{X}\\\;\HOLConst{INSERT}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}}.
% This completes the proof.
% \end{proof}

@@

There are two natural ways to implement
the \multivariate substitution $E\{\til P/\til X\}$, which replaces each free
occurrence of the variables $X_i$ in $E$ with the corresponding
$P_i$: (1) iterately applying the existing \univariate
\HOLinline{\HOLConst{CCS_Subst}}; (2) defining a new \multivariate substitution
function which substitutes all variables $\til X$ in parallel.
Note that there is the
possibility that $\til P$ syntactically again contains variables from $\til X$, thus
different orders of iterated substitutions may lead to different results.
To obtain the maximal flexibility with reduced efforts, we defined
a \multivariate version of \HOLinline{\HOLConst{CCS_Subst}}
called \HOLinline{\HOLConst{CCS_SUBST}}, based on HOL
\texttt{finite\_map} theory~\cite{holdesc}.

@@

The function \HOLinline{\HOLConst{CCS_SUBST}} takes a finite
map \HOLinline{\HOLFreeVar{fm}} (of the type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}) and a CCS expression,
and returns another CCS expression in which
all occurrences of variables in the finite domain of \HOLinline{\HOLFreeVar{fm}} are substituted with
the corresponding value in \HOLinline{\HOLFreeVar{fm}}. Initially, such a finite map can be built from
a list of variables \HOLinline{\HOLFreeVar{Xs}} and the corresponding substituted terms
\HOLinline{\HOLFreeVar{Ps}} by a helper function \HOLinline{\HOLConst{fromList}} (whose details are
omitted here).
\HOLinline{\HOLConst{CCS_SUBST}\\\;\ensuremath{(}\HOLConst{fromList}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Ps}\ensuremath{)}\\\;\HOLFreeVar{E}} is usually
abbreviated as \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}}.
Using \hl{finite} maps, the substitution mechanism is order-independent. For most CCS operators,
\HOLinline{\HOLConst{CCS_SUBST}}
 recursively calls itself on subterms. The most
interesting cases are at agent variables and recursion:
\begin{alltt}
   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{X} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{X}\hfill{[CCS_SUBST_var]}

@@

On the other hand, not every term within the CCS syntax is a context, as
context (or equation) variables are not allowed to occur within recursion. The converse
holds however: if the set of free variables of a CCS term
is disjoint from a list of variables, then the term is indeed a context
w.r.t. such list of variables:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{FV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}} \HOLConst{context} \HOLFreeVar{Xs} \HOLFreeVar{E}\hfill{[disjoint_imp_context]}
\end{alltt}
On the other hand, for any context (w.r.t.~$\til X$) of the form $\recu Y E$,
the set of free variables of $E$ excluding $Y$ is disjoint with $\til X$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{context} \HOLFreeVar{Xs} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{Y} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{FV} \HOLFreeVar{E} \HOLConst{DELETE} \HOLFreeVar{Y}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\hfill{[context_rec]}
\end{alltt}
Note that, in the above lemma we cannot
conclude \HOLinline{\HOLConst{DISJOINT}\\\;\ensuremath{(}\HOLConst{FV}\\\;\HOLFreeVar{E}\ensuremath{)}\\\;\ensuremath{(}\HOLConst{set}\\\;\HOLFreeVar{Xs}\ensuremath{)}},
because $Y$ may appear in both sets.
We also cannot conclude \HOLinline{\HOLConst{context}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{E}}, because in $\recu Y E$
the bound variable $Y$ may occur freely within another nested
recursion. For instance, variable $Y$ is free
in $\recu Y E$ , for $E = \recu Z (a.Y + b.Z) + c.Y$.

>>>>

In the literature (\citep[p.~102]{Gorrieri:2015jt}, e.g.), the variables $X_i$ of
a system of equations $\{X_i = E_i\}_{i\in I}$ are usually considered as
\emph{process variables} outside the CCS syntax.
Then, an equation body  $E_i$  is an \emph{open} expression
built with CCS operators plus these equation variables.
In a  formalisation, this means either defining a
whole new datatype (with the new equation variables), in which each CCS
operator must be duplicated, or adding equation variables as a
new primitive to the existing CCS type.
In either cases the \emph{disjointness} between equation variables and
agent variables is syntactically
guaranteed. Both solutions are rather combersome, and require a non-trivial modification
of the existing \univariate formalisation.

@@

As mentioned at the beginning of Section~\ref{s:eq},
within our CCS syntax an agent variable may occur outside the recursion in which it is bound.
Thus the same variable can appear both \emph{free} and \emph{bound} in a CCS term.
%
We denote the set of bound variables of a
given CCS term $E$ (the variables that are bound in a recursion
subexpression of $E$) as $\bv{E}$ (or \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}} in HOL), and the set of  free
variables as $\fv{E}$ (or \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}} in HOL). Both \HOLinline{\HOLConst{BV}} and \HOLinline{\HOLConst{FV}} have the
type ``\HOLinline{\ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS} \HOLTokenTransEnd \ensuremath{\alpha} \HOLTokenTransEnd \HOLTyOp{bool}}'', i.e. functions taking CCS terms and returning
(finite) sets of variables (of the type \HOLinline{\ensuremath{\alpha}}).
For their definition the interesting cases are those of
recursion and agent variables, shown below  (here \texttt{DELETE}
and \texttt{INSERT} are set-theoretic operators of HOL's
\texttt{pred_set} theory):
\begin{center}
\begin{tabular}{|l|l|}
\hline
\HOLConst{FV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLTokenLeftbrace{}\HOLFreeVar{X}\HOLTokenRightbrace{} & \HOLConst{FV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLTokenDefEquality{} \HOLConst{FV} \HOLFreeVar{p} \HOLConst{DELETE} \HOLFreeVar{X} \\
\HOLConst{BV} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLSymConst{\HOLTokenEmpty{}} & \HOLConst{BV} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{p}\ensuremath{)} \HOLTokenDefEquality{} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{p} \\
\hline
\end{tabular}
\end{center}
Furthermore, $E$ is a process, written \HOLinline{\HOLConst{IS_PROC} \HOLFreeVar{E}}, if
it does not contain any free variable, i.e. $\fv{E} = \emptyset$:
\begin{alltt}
   \HOLConst{IS_PROC} \HOLFreeVar{E} \HOLTokenDefEquality{} \HOLConst{FV} \HOLFreeVar{E} \HOLSymConst{\ensuremath{=}} \HOLSymConst{\HOLTokenEmpty{}}\hfill{[IS_PROC_def]}
\end{alltt}
And a list of CCS processes can be asserted by \HOLinline{\HOLConst{ALL_PROC}} defined upon \HOLinline{\HOLConst{IS_PROC}}:
\begin{alltt}
   \HOLConst{ALL_PROC} \HOLFreeVar{Es} \HOLTokenDefEquality{} \HOLConst{EVERY} \HOLConst{IS_PROC} \HOLFreeVar{Es}\hfill{[ALL_PROC_def]}
\end{alltt}

The set of free and bound
variables of a term need not be disjoint, e.g.~in $X + \recu X E$.
More importantly, when going
from a CCS expression to its sub-expressions, the set of free
variables may increase, while the set of bound variables either
remains the same or decreases. For instance, $\fv{\recu X (\mu.X)} = \emptyset$, while
$\fv{\mu.X} = \{X\}$. This property of $\fv{\cdot}$ brings some
%% "transition induction" (1/2)
difficulities in transition inductions. As an evidence,
we prove the following fundamental property of
$\fv{\cdot}$~\citep[p.~1209]{milner1990operational}:
%
\begin{proposition}
\label{prop:transFV}
The derivatives of a process are themselves processes, i.e.
if $E \arr{\mu} E'$ and $\fv{E} = \emptyset$, then $\fv{E'} =
\emptyset$. Formally:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{IS_PROC} \HOLBoundVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLConst{IS_PROC} \ensuremath{\HOLBoundVar{E}\sp{\prime}}\hfill{[TRANS_PROC]}
\end{alltt}
\end{proposition}

@@

As a matter of fact, in the HOL proof ``action induction'' (also
known as \emph{transition induction})

@@

Note that \HOLinline{\HOLConst{FV}\\\;\HOLFreeVar{E}\\\;\HOLConst{DELETE}\\\;\HOLFreeVar{X}\\\;\HOLSymConst{\ensuremath{=}}\\\;\HOLConst{FV}\\\;\ensuremath{(}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{)}}, thus
the current proof goal is indeed the consequence of action induction on the
recursion operator.

@@

Proposition~\ref{prop:transFV} is essential in the proofs of the
  \multivariate versions of all unique-solution theorems.
On the other hand, the analogous
result for bound variables is indeed just a simple
%% transition induction
transition induction. (The easy proof is omitted.)
\begin{proposition}
\label{prop:transBV}
if $E \arr{\mu} E'$ then $\bv{E'} \subseteq \bv{E}$, or formally:
\begin{alltt}
\HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{E} \HOLBoundVar{u} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLBoundVar{E} \HOLTokenTransBegin\HOLBoundVar{u}\HOLTokenTransEnd \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{BV} \ensuremath{\HOLBoundVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLBoundVar{E}\hfill{[TRANS_BV]}
\end{alltt}
\end{proposition}
% \begin{proof}
% Following the proof of Proposition~\ref{prop:transFV}, we do action
% induction on $E \arr{\mu} E'$ and simplify all subgoals by set-theoretic
% facts. The only subgoal left is:
% \begin{alltt}
%         \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{E}}
%    ------------------------------------
%     0.  \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}}
% \end{alltt}
% Note the difference with the proof of Proposition~\ref{prop:transFV}: now
% we have \texttt{INSERT} in place of \texttt{DELETE}. Then, we directly
% use the assumption and the transitivity of $\subseteq$ and try to
% prove the following goal instead:
% \begin{alltt}
%         \HOLinline{\HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLFreeVar{X} \HOLConst{INSERT} \HOLConst{BV} \HOLFreeVar{E}}
%    ------------------------------------
%     0.  \HOLinline{\HOLConst{BV} \ensuremath{\HOLFreeVar{E}\sp{\prime}} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \ensuremath{(}\ensuremath{[}\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]} \HOLFreeVar{E}\ensuremath{)}}
% \end{alltt}
% This is an easy goal, due to the following lemma which is
%  proved by doing induction on $E$:
% \begin{alltt}
% \HOLTokenTurnstile{} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{X} \HOLBoundVar{E} \ensuremath{\HOLBoundVar{E}\sp{\prime}}. \HOLConst{BV} \ensuremath{(}\ensuremath{[}\ensuremath{\HOLBoundVar{E}\sp{\prime}}\ensuremath{/}\HOLBoundVar{X}\ensuremath{]} \HOLBoundVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenSubset{}} \HOLConst{BV} \HOLBoundVar{E} \HOLSymConst{\HOLTokenUnion{}} \HOLConst{BV} \ensuremath{\HOLBoundVar{E}\sp{\prime}}\hfill{[BV_SUBSET]}
% \end{alltt}
% Taking $E ' = \recu X E$ in the above lemma, we have
%     \HOLinline{\HOLConst{BV}\\\;\ensuremath{(}\ensuremath{[}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{/}\HOLFreeVar{X}\ensuremath{]}\\\;\HOLFreeVar{E}\ensuremath{)}\\\;\HOLSymConst{\HOLTokenSubset{}}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}\\\;\HOLSymConst{\HOLTokenUnion{}}\\\;\HOLConst{BV}\\\;\ensuremath{(}\HOLConst{rec}\\\;\HOLFreeVar{X}\\\;\HOLFreeVar{E}\ensuremath{)}},
%     the right side equals to \HOLinline{\HOLConst{BV}\\\;\HOLFreeVar{E}\\\;\HOLSymConst{\HOLTokenUnion{}}\\\;\ensuremath{(}\HOLFreeVar{X}\\\;\HOLConst{INSERT}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}\ensuremath{)}} $=$
%     \HOLinline{\HOLFreeVar{X}\\\;\HOLConst{INSERT}\\\;\HOLConst{BV}\\\;\HOLFreeVar{E}}.
% This completes the proof.
% \end{proof}

@@

There are two natural ways to implement
the \multivariate substitution $E\{\til P/\til X\}$, which replaces each free
occurrence of the variables $X_i$ in $E$ with the corresponding
$P_i$: (1) iterately applying the existing \univariate
\HOLinline{\HOLConst{CCS_Subst}}; (2) defining a new \multivariate substitution
function which substitutes all variables $\til X$ in parallel.
Note that there is the
possibility that $\til P$ syntactically again contains variables from $\til X$, thus
different orders of iterated substitutions may lead to different results.
We thus prefer to define a \multivariate version of \HOLinline{\HOLConst{CCS_Subst}}
called \HOLinline{\HOLConst{CCS_SUBST}}, based on HOL
\texttt{finite\_map} theory~\cite{holdesc}.

@@

The function \HOLinline{\HOLConst{CCS_SUBST}} takes a finite
map \HOLinline{\HOLFreeVar{fm}} (of the type \HOLinline{\ensuremath{\alpha} \HOLTokenMapto{} \ensuremath{(}\ensuremath{\alpha}, \ensuremath{\beta}\ensuremath{)} \HOLTyOp{CCS}}) and a CCS expression,
and returns another CCS expression in which
all occurrences of variables in the finite domain of \HOLinline{\HOLFreeVar{fm}} are substituted with
the corresponding value in \HOLinline{\HOLFreeVar{fm}}. Initially, such a finite map can be built from
a list of variables \HOLinline{\HOLFreeVar{Xs}} and the corresponding substituted terms
\HOLinline{\HOLFreeVar{Ps}} by a helper function \HOLinline{\HOLConst{fromList}} (whose details are
omitted here).
\HOLinline{\HOLConst{CCS_SUBST}\\\;\ensuremath{(}\HOLConst{fromList}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{Ps}\ensuremath{)}\\\;\HOLFreeVar{E}} is usually
abbreviated as \HOLinline{\ensuremath{[}\HOLFreeVar{Ps}\ensuremath{/}\HOLFreeVar{Xs}\ensuremath{]}\\\;\HOLFreeVar{E}}.
Using finite maps, the substitution mechanism is order-independent. For most CCS operators,
\HOLinline{\HOLConst{CCS_SUBST}}
 recursively calls itself on subterms. The most
interesting cases are at agent variables and recursion:
\begin{alltt}
   \HOLConst{CCS_SUBST} \HOLFreeVar{fm} \ensuremath{(}\HOLConst{var} \HOLFreeVar{X}\ensuremath{)} \HOLTokenDefEquality{} \HOLKeyword{if} \HOLFreeVar{X} \HOLSymConst{\HOLTokenIn{}} \HOLConst{FDOM} \HOLFreeVar{fm} \HOLKeyword{then} \HOLFreeVar{fm} \HOLConst{'} \HOLFreeVar{X} \HOLKeyword{else} \HOLConst{var} \HOLFreeVar{X}\hfill{[CCS_SUBST_var]}

@@

Not every term within the CCS syntax is a context, as
context (or equation) variables are not allowed to occur within recursion. The converse
holds however: if the set of free variables of a CCS term
is disjoint from a list of variables, then the term is indeed a context
w.r.t. such list of variables:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{FV} \HOLFreeVar{E}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}} \HOLConst{context} \HOLFreeVar{Xs} \HOLFreeVar{E}\hfill{[disjoint_imp_context]}
\end{alltt}
On the other hand, for any context (w.r.t.~$\til X$) of the form $\recu Y E$,
the set of free variables of $E$ excluding $Y$ is disjoint with $\til X$:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{context} \HOLFreeVar{Xs} \ensuremath{(}\HOLConst{rec} \HOLFreeVar{Y} \HOLFreeVar{E}\ensuremath{)} \HOLSymConst{\HOLTokenImp{}} \HOLConst{DISJOINT} \ensuremath{(}\HOLConst{FV} \HOLFreeVar{E} \HOLConst{DELETE} \HOLFreeVar{Y}\ensuremath{)} \ensuremath{(}\HOLConst{set} \HOLFreeVar{Xs}\ensuremath{)}\hfill{[context_rec]}
\end{alltt}
In the above lemma we cannot
conclude \HOLinline{\HOLConst{DISJOINT}\\\;\ensuremath{(}\HOLConst{FV}\\\;\HOLFreeVar{E}\ensuremath{)}\\\;\ensuremath{(}\HOLConst{set}\\\;\HOLFreeVar{Xs}\ensuremath{)}},
because $Y$ may appear in both sets.
We also cannot conclude \HOLinline{\HOLConst{context}\\\;\HOLFreeVar{Xs}\\\;\HOLFreeVar{E}}, because in $\recu Y E$
the bound variable $Y$ may occur freely within another nested
recursion. For instance, variable $Y$ is free
in $\recu Y E$ , for $E = \recu Z (a.Y + b.Z) + c.Y$.
